---
title: "miRNA Differential expression analysis for all plasma & serum samples"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
# All plasma & serum samples
```{python3}
import pandas as pd
samples = pd.read_csv('data/sample_sheet.csv')
# must be Healthy Control Study and must pass MiSeq QC
samples = samples.loc[(samples['Study'] == 'Healthy Controls') & (samples['MISEQ.QC.PASS'] == 'PASS')]  
samples = samples.set_index('MT.Unique.ID').sort_values(by=['Participant.ID', 'Source'])
# capitalize & strip whitespace for consistency
for column in ['Gender', 'Race', 'Source']:
    samples[column] = samples[column].str.capitalize()
    samples[column] = samples[column].str.strip()
# use correct ontology terms
race_ontology = {'Asian': 'Asian',
 'Black or african american': 'African_American',
 'Mixed/asian & white': 'Multiracial',
 'Mixed/asian &black': 'Multiracial',
 'Mixed/black, white, asian': 'Multiracial',
 'Native hawiian or other pacific islander': 'Pacific_Islander',
 'Pacific islander': 'Pacific_Islander',
 'White': 'White'}
for id in samples.index:
    race = samples.at[id, 'Race']
    samples.at[id, 'Race'] = race_ontology[race] if race in race_ontology else 'Multiracial'
samples.index = pd.Index(['X' + str(row) for row in samples.index])
samples.to_csv('data/samples.csv')
```
## Load the sample data and miRNA counts
```{r}
samples <- read.csv('data/samples.csv')
samples <- subset(samples, Library.Generation.Set != "SetRecheck" )  # exclude SetRecheck samples
samples <- samples[samples$X != "X11" & samples$X != "X92",]  # remove outliers
counts <- read.csv('data/get_canonical/canon_mir_counts.csv')
samples$Participant.ID <- factor(samples$Participant.ID)  # ID is categorical, not numerical
samples$Sample.ID <- factor(samples$Sample.ID)
rownames(counts) = counts$X
rownames(samples) = samples$X
counts$X <- NULL  # remove extra column
samples$X <- NULL
counts <- counts[,rownames(samples)]
head(samples)
head(counts)
plasma_samples <- subset(samples, Source=="Plasma")
serum_samples <- subset(samples, Source=="Serum")
plasma_samples$MT.Unique.ID <- rownames(plasma_samples)
serum_samples$MT.Unique.ID <- rownames(serum_samples)
plasma_counts <- counts[rownames(plasma_samples)]
serum_counts <- counts[rownames(serum_samples)]
```
# Plot consistency of detected miRNAs (Fig 2C)
```{r}
library(ggplot2)
# set ggplot theme
my_theme_bw <- theme_bw()
theme_set(my_theme_bw)
my_theme_bw <- theme_update(text=element_text(color="black"), axis.text=element_text(color="black", size=rel(1.0)), axis.title=element_text(color="black"), strip.text=element_text(color="black"))
# separate data for plasma & serum
count_plasma_samples <- length(rownames(plasma_samples))
#plasma_counts <- plasma_counts[which(rowSums(plasma_counts > 0) > 0),]
plasma_miRNAs <- data.frame(sample_fraction = (rowSums(plasma_counts > 0) / count_plasma_samples))
plasma_miRNAs$Source <- "Plasma"
count_serum_samples <- length(rownames(serum_samples))
#serum_counts <- serum_counts[which(rowSums(serum_counts > 0) > 0),]
serum_miRNAs <- data.frame(sample_fraction = (rowSums(serum_counts > 0) / count_serum_samples))
serum_miRNAs$Source <- "Serum"
detected_miRNAs <- rbind(plasma_miRNAs, serum_miRNAs)
ggplot(detected_miRNAs, aes(sample_fraction, fill=Source)) + geom_histogram(alpha=0.5, position='identity', bins=30) + ggtitle("miRNA detection") + xlab("Fraction of Samples") + ylab("Number of miRNAs detected")
```
## Plot library size distribution for plasma and serum library samples (total reads mapped to miRs)
```{r}
plasma_library_sizes <- data.frame(reads=colSums(plasma_counts))
plasma_library_sizes$sample <- rownames(plasma_library_sizes)
plasma_library_sizes$source <- "Plasma"
ggplot(plasma_library_sizes, aes(reads)) + geom_histogram(alpha=0.5) + ggtitle("Library Size Distribution in Plasma") + xlab("Read Count") + ylab("Sample Count")

serum_library_sizes <-data.frame(reads=colSums(serum_counts))
serum_library_sizes$sample <- rownames(serum_library_sizes)
serum_library_sizes$source <- "Serum"
ggplot(serum_library_sizes, aes(reads)) + geom_histogram(alpha=0.5) + ggtitle("Library Size Distribution in Serum") + xlab("Read Count") + ylab("Sample Count")

library_sizes_df <- rbind(serum_library_sizes, plasma_library_sizes)
ggplot(library_sizes_df, aes(reads, fill=Source))  + geom_histogram(alpha=0.5, position='identity', bins=30) + ggtitle("Library Size Distribution") + xlab("Read Count") + ylab("Sample Count")
```
## Plasma Downsampling for Figure 2C to control for biases in different library sizes
```{r}
library(data.table)
library(vegan)
# DOWNSAMPLING: Plasma
downsample.to.vec <- c(Inf, round(10^seq(4, 5.5, 0.5, 6), 0))
drr.list.plasma <- sapply(downsample.to.vec, simplify=FALSE, function(x) t(drarefy(t(plasma_counts), x)))

names(drr.list.plasma) <- downsample.to.vec
drr.all.plasma <- data.table(rbindlist(lapply(drr.list.plasma, melt), idcol=TRUE, use.names=TRUE), keep.rownames=TRUE)
setnames(drr.all.plasma, c("downsample.to", "miR.id", "MT.Unique.ID", "est.prob.detected"))
plasma_samples$lib.size <- NA
for(mt_unique_id in rownames(plasma_samples)) {
  plasma_samples[mt_unique_id, "lib.size"] <- sum(plasma_counts[mt_unique_id])
}
drr.all.plasma.merge <- merge(drr.all.plasma, plasma_samples)
drr.all.plasma.merge[, downsample.to:=as.numeric(downsample.to)]
drr.all.plasma.filt <- subset(drr.all.plasma.merge, lib.size>=downsample.to|lib.size==Inf)

# plot cutoff detected
min.prob.detected=0.90
est.prob.detected.n.miRs.plasma <- drr.all.plasma.filt[,
                                               .(n.detected=sum(ifelse(est.prob.detected>min.prob.detected, 1, 0)),
                                                  n.not.detected=sum(ifelse(est.prob.detected<=min.prob.detected, 1, 0)),
                                                  n.total=.N),
                                                by=.(MT.Unique.ID,downsample.to,Source)]
# Plasma miRs detected
est.prob.detected.n.miRs.plasma$downsample.to.char <- paste("10^", as.character(round(log10(est.prob.detected.n.miRs.plasma$downsample.to), 1)), sep='')
ggplot(est.prob.detected.n.miRs.plasma, aes( x=downsample.to.char, y=n.detected)) + 
  geom_boxplot(aes(group=downsample.to)) + ggtitle("Library Downsampling") + ylab('miRNAs Detected') + xlab('Library Downsample Size')
```

# Filter
```{r}
library(edgeR)
design <- model.matrix(~Source, samples)
dge = DGEList(counts = counts, samples = samples)
# require miRNAs to have CPM > 1 in at least 2 samples
countsPerMillion <- edgeR::cpm(dge)
countCheck <- countsPerMillion > 1
head(countCheck)
keep <- which(rowSums(countCheck) >= 2) 
dge <- dge[keep,]
```
# Explore variance
```{r}
library(SingleCellExperiment)
library(scater)
reads_sce <- SingleCellExperiment(assays=list(counts=dge$counts),  colData=dge$samples)
# remove unexpressed miRNAs
keep_feature <- rowSums(counts(reads_sce) > 0) > 0
reads_sce <- reads_sce[keep_feature, ]
reads_sce <- calculateQCMetrics(reads_sce)
# create separate plasma & serum SCEs & DGEs
#plasma_samples <- subset(colData(reads_sce), Source == "Plasma")
#plasma_samples <- as.data.frame(plasma_samples)
#plasma_samples$MT.Unique.ID <- rownames(plasma_samples)
#plasma_counts <- as.data.frame(counts(reads_sce))[rownames(plasma_samples)]
plasma_sce <- SingleCellExperiment(assays=list(counts=as.matrix(plasma_counts)), colData=plasma_samples)
plasma_sce <- calculateQCMetrics(plasma_sce)
#serum_samples <- subset(colData(reads_sce), Source == "Serum")
#serum_counts <- as.data.frame(counts(reads_sce))[rownames(serum_samples)]
serum_sce <- SingleCellExperiment(assays=list(counts=as.matrix(serum_counts)), colData=serum_samples)
serum_sce <- calculateQCMetrics(serum_sce)
# log transform
cpm(reads_sce) <- calculateCPM(reads_sce)
reads_sce <- normalize(reads_sce)
logcounts(reads_sce) <- log2(calculateCPM(reads_sce) + 1)
# visualize
hist(reads_sce$total_counts, breaks=100)  # counts per sample
hist(reads_sce$total_features, breaks=100)  # counts per miRNA
plotQC(reads_sce, type = "highest-expression") + ggtitle("Plasma & Serum")
plotQC(plasma_sce, type = "highest-expression") + ggtitle("Plasma")
plotQC(serum_sce, type = "highest-expression") + ggtitle("Serum")
plotQC(reads_sce, type="explanatory-variables", variables=c("Index", "Participant.ID", "Collection.Date", "Library.Generation.Set", "MiSeq.QC.Run", "Source", 'total_features', "Age", "Race", "Gender"))
```



## Examine sources of variation
```{r}
plotPCA(reads_sce, exprs_values = "logcounts", colour_by = "Library.Generation.Set", size_by = "total_features", shape_by = "Source")
for (var in c("total_features", "Age", "Library.Generation.Set", "Index", "Participant.ID", "Source", "Race", "Gender", "Collection.Date")) {
  print(
    plotQC(reads_sce, type = "find-pcs", exprs_values = "logcounts", variable = var)
    )  + ggtitle(var) 
  }
```

## Normalize & Remove unwanted sources of variation
```{r}
library(RUVSeq)
library(mvoutlier)
dge <- calcNormFactors(dge)
dge <- estimateGLMCommonDisp(dge, design)
dge <- estimateGLMTagwiseDisp(dge, design)
fit <- glmFit(dge, design)
res <- residuals(fit, type="deviance")
set <- newSeqExpressionSet(dge$counts, phenoData=dge$samples)
ruvr_sets <- list()
for(k in 1:5) {
  ruvr_sets[[k]] <- RUVr(set, row.names(dge), k=k, res)
  assay(reads_sce, paste("RUVr k=", toString(k))) <- log2(t(t(assayData(ruvr_sets[[k]])$normalizedCounts) / colSums(assayData(ruvr_sets[[k]])$normalizedCounts) * 1e6) + 1)
}
for(n in assayNames(reads_sce)) {
  print(
        plotPCA(
            reads_sce,
            colour_by = "Library.Generation.Set",
            size_by = "total_features",
            shape_by = "Source",
            exprs_values = n
        ) + ggtitle(n)
  )
}
```
## Detect outliers
```{r}
reads_sce <- runPCA(reads_sce, use_coldata = TRUE, detect_outliers = TRUE)
outliers <- colnames(reads_sce)[reads_sce$outlier]
head(outliers)
```
## Examine sources of variance after removing unwanted variation
```{r}
for (var in c("total_features", "Age", "Library.Generation.Set", "Index", "Participant.ID", "Source", "Race", "Gender", "Collection.Date")) {
  print(
    plotQC(reads_sce, type = "find-pcs", exprs_values = "RUVr k= 2", variable = var)
    )  + ggtitle(var) 
}
```
# Visualize top highly-expressed plasma-vs-serum miRNAs
```{r}
library(RColorBrewer)
library(reshape2)
# Ryan's code, modified
norm.expr.matr <- exprs(reads_sce)
# Rank the mean expression values for plasma/serum miRs. Highest expression = 1
mean.expr.plasma.rank <- rank(-1*rowMeans(norm.expr.matr[, colData(reads_sce)$Source=="Plasma"]))
mean.expr.serum.rank <- rank(-1*rowMeans(norm.expr.matr[, colData(reads_sce)$Source=="Serum"]))
top_N <- 18  # top_N=18 resuls in 20 miRNAs in the plot
top.miRs <- row.names(norm.expr.matr)[mean.expr.plasma.rank <= top_N | mean.expr.serum.rank <= top_N] # get the names of the top miRs in plasma or serum
norm.expr.top <- norm.expr.matr[top.miRs, ] # Get the expression matrix for the top mIRs
norm.expr.melt <- reshape2::melt(norm.expr.top) # Convert your normalized expression matrix to a 3 column data.frame (row.name, col.name, expression value). Melt is in the dpylr package, I believe.
colnames(norm.expr.melt) <- c("miR.ID", "MT.Unique.ID", "norm.expr") # just so it's easier for me to tell you which columns I'm using.
norm.expr.melt$Source <- ""
for (row_num in 1:nrow(norm.expr.melt)){  # Pull the plasma/serum source value from the column metadata.
  mt_unique_id <- norm.expr.melt[row_num, ]$MT.Unique.ID
  norm.expr.melt[row_num, "Source"] <- as.character(samples[mt_unique_id,"Source"])
}
# This would be a simple boxplot with plasma/serum side-by-side. Overlaying the boxes over points takes a little more tweaking to get the dodge/width right, but it's doable.
ggplot(norm.expr.melt, aes(x=reorder(miR.ID, norm.expr, FUN=median), y=norm.expr, fill=Source)) + geom_boxplot(pos="dodge", outlier.size=0.5) + 
  ggtitle("Top 20 Expressed miRNAs") + ylab("Normalized Expression") + xlab("miRNA ID") + 
  theme(panel.grid.major.x = element_line(size = 0.5, linetype = 'solid', colour = "grey"), 
        panel.grid.major.y = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1))
ggplot(norm.expr.melt, aes(x=reorder(miR.ID, norm.expr, FUN=median), y=norm.expr, fill=Source)) + geom_boxplot(pos="dodge", outlier.size=0.5) + 
  ggtitle("Top 20 Expressed miRNAs") + ylab("Normalized Expression") + xlab("miRNA ID") + 
  theme(panel.grid.major.y = element_line(size = 0.5, linetype = 'solid', colour = "grey"), 
        panel.grid.major.x = element_blank()) + coord_flip()
```

# DE analysis with EdgeR
```{r}
library(pheatmap)
# based on Ryan's code, pass RUVSeq-corrected data to edgeR
ruvr2 <- ruvr_sets[[2]]
design <- model.matrix(~Source + W_1 + W_2, pData(ruvr2))
dge <- estimateDisp(dge, design = design, tagwise = TRUE, robust = TRUE)
fit <- glmFit(dge, design)
# contrast source (plasma vs serum)
lrt <- glmLRT(fit, coef="SourceSerum")
results <- data.frame(topTags(lrt, n=Inf, sort.by="PValue", p.value=0.05))
sig_miRs = list()
alpha_level <- 0.05
for (logFC_threshold in 1:2) {
  sig_miRs[[logFC_threshold]] <- rownames(results[results$PValue < alpha_level & results$FDR < alpha_level & abs(results$logFC) > logFC_threshold,])  # filter by p-value and logFC
}
sig_miRs[[3]] <- rownames(results[results$PValue < 0.05,][0:50,])
# heatmaps of significant DE miRNAs
Source <- pData(ruvr2)[,c("Source")]
annotations <- as.data.frame(Source)
rownames(annotations) <- rownames(pData(ruvr2))
for(miR_list in sig_miRs) {
  pheatmap(norm.expr.matr[miR_list,], cluster_rows=TRUE, show_rownames=TRUE, cluster_cols=TRUE, annotation_col=annotations, main="Differentially Expressed miRNAs")
}
```

```{r}
save.image("diff_expr_all_plasma_vs_serum.RData")
```

