---
title: "Differential expression analysis using DESeq2 for matched plasma & serum samples"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
# Match plasma & serum samples
```{python3}
import pandas as pd
samples = pd.read_csv('data/sample_sheet.csv')
# must be Healthy Control Study and must pass MiSeq QC
samples = samples.loc[(samples['Study'] == 'Healthy Controls') & (samples['MISEQ.QC.PASS'] == 'PASS')]  
samples = samples.set_index('MT.Unique.ID').sort_values(by=['Participant.ID', 'Source'])
# capitalize & strip whitespace for consistency
for column in ['Gender', 'Race', 'Source']:
    samples[column] = samples[column].str.capitalize()
    samples[column] = samples[column].str.strip()
# use correct ontology terms
race_ontology = {'Asian': 'Asian',
 'Black or african american': 'African_American',
 'Mixed/asian & white': 'Multiracial',
 'Mixed/asian &black': 'Multiracial',
 'Mixed/black, white, asian': 'Multiracial',
 'Native hawiian or other pacific islander': 'Pacific_Islander',
 'Pacific islander': 'Pacific_Islander',
 'White': 'White'}
for id in samples.index:
    race = samples.at[id, 'Race']
    samples.at[id, 'Race'] = race_ontology[race] if race in race_ontology else 'Multiracial'
# get matched plasma & serum samples
mir_counts = pd.read_csv("data/get_canonical/canon_mir_counts.csv", index_col=0)
samples.index = pd.Index(['X' + str(row) for row in samples.index])
serum_part_ids = set(samples.loc[samples['Source'] =='Serum']['Participant.ID'])
matched_samples = samples.loc[samples['Participant.ID'].isin(serum_part_ids)]
matched_mir_counts = mir_counts[matched_samples.index]
matched_samples.to_csv('data/matched_plasma-serum_samples.csv')
matched_mir_counts.to_csv('data/matched_plasma-serum_mir_counts.csv')
```
## Load the sample data and miRNA counts
```{r}
samples <- read.csv('data/matched_plasma-serum_samples.csv')
counts <- read.csv('data/matched_plasma-serum_mir_counts.csv')
samples$Participant.ID <- factor(samples$Participant.ID)  # ID is categorical, not numerical
samples$Sample.ID <- factor(samples$Sample.ID)
rownames(counts) = counts$X
rownames(samples) = samples$X
counts$X <- NULL  # remove extra column
counts <- counts[,rownames(samples)]
head(samples)
head(counts)
plasma_samples <- subset(samples, Source == "Plasma")
plasma_counts <- counts[rownames(plasma_samples)]
serum_samples <- subset(samples, Source == "Serum")
serum_counts <- counts[rownames(serum_samples)]
```

# Explore variance & Filter
```{r}
library(SingleCellExperiment)
library(scater)
reads_sce <- SingleCellExperiment(assays=list(counts=as.matrix(counts)),  colData=samples)
plasma_sce <- SingleCellExperiment(assays=list(counts=as.matrix(plasma_counts)), colData=plasma_samples)
serum_sce <- SingleCellExperiment(assays=list(counts=as.matrix(serum_counts)), colData=serum_samples)
# remove unexpressed miRNAs
keep_feature <- rowSums(counts(reads_sce) > 0) > 0
reads_sce <- reads_sce[keep_feature, ]
reads_sce <- calculateQCMetrics(reads_sce)
plasma_sce <- plasma_sce[rowSums(counts(plasma_sce)>0)>0,]
plasma_sce <- calculateQCMetrics(plasma_sce)
serum_sce <- serum_sce[rowSums(counts(plasma_sce)>0)>0,]
serum_sce <- calculateQCMetrics(serum_sce)
# log transform
cpm(reads_sce) <- calculateCPM(reads_sce)
reads_sce <- normalize(reads_sce)
logcounts(reads_sce) <- log2(calculateCPM(reads_sce) + 1)
# visualize
hist(reads_sce$total_counts, breaks=100)  # counts per sample
hist(reads_sce$total_features, breaks=100)  # counts per miRNA
plotQC(reads_sce, type = "highest-expression")
plotQC(plasma_sce, type = "highest-expression") + ggtitle("Plasma")
plotQC(serum_sce, type = "highest-expression") + ggtitle("Serum")
plotQC(reads_sce, type="explanatory-variables", variables=c("Index", "Participant.ID", "Collection.Date", "Library.Generation.Set", "MiSeq.QC.Run", "Source", 'total_features'))
```
## Examine sources of variance
```{r}
plotPCA(reads_sce, exprs_values = "logcounts", colour_by = "Library.Generation.Set", size_by = "total_features", shape_by = "Source")
for (var in c("total_features", "Age", "Library.Generation.Set", "Index", "Participant.ID", "Source", "Race", "Gender", "Collection.Date")) {
  print(
    plotQC(reads_sce, type = "find-pcs", exprs_values = "logcounts", variable = var)
    )  + ggtitle(var) 
  }
```

## Remove unwanted sources of variation
```{r}
library(RUVSeq)
library(edgeR)
library(ggplot2)
design <- model.matrix(~0+Source+Race+Gender+Age, samples)
dge = DGEList(counts = counts(reads_sce), samples = colData(reads_sce))
dge <- calcNormFactors(dge)
dge <- estimateGLMCommonDisp(dge, design)
dge <- estimateGLMTagwiseDisp(dge, design)
fit <- glmFit(dge, design)
res <- residuals(fit, type="deviance")
set <- newSeqExpressionSet(dge$counts, phenoData=dge$samples)
ruvr1 <- RUVr(set, row.names(dge), k=1, res)
assay(reads_sce, "RUVr1") <- log2(t(t(assayData(ruvr1)$normalizedCounts) / colSums(assayData(ruvr1)$normalizedCounts) * 1e6) + 1)
for(n in assayNames(reads_sce)) {
  print(
        plotPCA(
            reads_sce,
            colour_by = "Library.Generation.Set",
            size_by = "total_features",
            shape_by = "Source",
            exprs_values = n
        ) + ggtitle(n)
  )
}
```
## Examine sources of variance after removing unwanted variation
```{r}
for (var in c("total_features", "Age", "Library.Generation.Set", "Index", "Participant.ID", "Source", "Race", "Gender", "Collection.Date")) {
  print(
    plotQC(reads_sce, type = "find-pcs", exprs_values = "RUVr1", variable = var)
    )  + ggtitle(var) 
}
```

# Run the DESeq2 pipeline
```{r}
library(DESeq2)  # TODO: use filtered & RUVSeq-corrected data
deseq2Data <- DESeqDataSetFromMatrix(countData = counts, design = ~ Participant.ID + Source, colData = samples)
deseq2Data <- DESeq(deseq2Data)
results <- results(deseq2Data, cooksCutoff=FALSE, independentFiltering=FALSE)
head(as.data.frame(results))
```

# Visualize
## Create a heatmap of the variance-stabilizing-transformed expression data
```{r}
library(pheatmap)
vstData <- varianceStabilizingTransformation(deseq2Data, blind=FALSE)  # transform the data
select <- order(rowMeans(counts(deseq2Data,normalized=TRUE)), decreasing=TRUE)
Source <- colData(deseq2Data)[,c("Source")]
annotations <- as.data.frame(Source)
rownames(annotations) <- colnames(deseq2Data)
pheatmap(assay(vstData)[select,], cluster_rows=TRUE, show_rownames=FALSE, cluster_cols=TRUE, annotation_col=annotations)
```

## Create a heatmap of sample-to-sample distances
```{r}
library("RColorBrewer")
sampleDists <- dist(t(assay(vstData)))  # compute sample-to-sample distances
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vstData$Participant.ID, vstData$Source, sep=" - ")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

## Create a PCA plot
```{r}
pcaData <- plotPCA(vstData, intgroup=c("Source"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=Source)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```
```{r}
save.image("diff_expr_plasma_vs_serum.RData")
```

